---
title: "Choosing TTC x Provides the Most Time Stability during Rush Hours"
subtitle: "Analysis on the Delay Magnitude of different TTC Services"
author: 
  - Sandy Yu
thanks: "Code and data are available at: https://github.com/Jingying-yu/ttc_delay.git"
date: today
date-format: long
abstract: ""
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
# load librarys and datasets
library(tidyverse)
library(dplyr)
library(lubridate)
library(zoo)
library(knitr)
library(ggplot2)

bus_delay <- read.csv(here::here("inputs/data/cleaned_data/bus_delay.csv"))
streetcar_delay <- read.csv(here::here("inputs/data/cleaned_data/streetcar_delay.csv"))
subway_delay <- read.csv(here::here("inputs/data/cleaned_data/subway_delay.csv"))
subway_delay_code <- read.csv(here::here("inputs/data/cleaned_data/subway_delay_code.csv"))

```


# Introduction



Analyses and findings in this paper are structured into several sections: @sec-data -- Data, @sec-results -- Results, and @sec-discussion -- Discussion. The Data section introduces the context of the datasets and examines all datasets and variables kept for analysis. The Result section focuses on visualizing and presenting the data results in Data section. The Discussion section further evaluate the implications behind the data results presented in the previous section, expands the topic beyond pure data, and touches on any weaknesses and next steps.


# Data {#sec-data}




## TTC Bus Delay



```{r}
#| label: tbl-bus-delay
#| tbl-cap: Sample of Cleaned TTC Bus Delay Dataset
#| echo: false

sample_bus_delay <- head(bus_delay, n = 3)

knitr::kable(sample_bus_delay, col.names = c("Date", "Line", "Time", "Day", "Location", "Cause of Delay", "Min Delayed", "Min Gap", "Direction", "Vehicle"), align = c("l","c", "c", "c", "c","c", "c", "c", "c","c"))
```


## TTC Streetcar Delay


```{r}
#| label: tbl-streetcar-delay
#| tbl-cap: Sample of Cleaned TTC Streetcar Delay Dataset
#| echo: false

sample_streetcar_delay <- head(streetcar_delay, n = 3)

knitr::kable(sample_streetcar_delay, col.names = c("Date", "Line", "Time", "Day", "Location", "Cause of Delay", "Min Delayed", "Min Gap", "Bound", "Vehicle"), align = c("l","c", "c", "c", "c","c", "c", "c", "c","c"))
```


## TTC Subway Delay

### Delay Code
```{r}
#| label: tbl-subway-code
#| tbl-cap: Sample of TTC Subway Delay Code
#| echo: false
#| message: false

sample_subway_delay_code <- head(subway_delay_code, n = 3)

knitr::kable(sample_subway_delay_code, col.names = c("Code", "Code Description"), align = c("c","c"))
```

### Delay Data
```{r}
#| label: tbl-subway-delay
#| tbl-cap: Sample of Cleaned TTC Subway Delay Dataset
#| echo: false
#| message: false

sample_subway_delay <- head(subway_delay, n = 3)

knitr::kable(sample_subway_delay, col.names = c("Date", "Time", "Day", "Location", "Delay Code", "Min Delayed", "Min Gap", "Bound", "Line", "Vehicle"), align = c("l","c", "c", "c", "c","c", "c", "c", "c","c"))
```

## TTC Delay Comparision



# Results {#sec-result}



## Highest Number of Total Delayed

```{r}
#| label: fig-total-delay
#| fig-cap: Visualization to show the TTC service that has the highest number of delays in year 2023
#| echo: false
#| message: false
#construct appropriate table for visualization
service_type <- c("bus", "streetcar", "subway")
num_delay <- c(nrow(bus_delay), nrow(streetcar_delay), nrow(subway_delay))

delay_count <- data.frame(service_type, num_delay)

# Generate the bar plot
ggplot(delay_count, aes(x = service_type, y = num_delay, fill = service_type)) +
  geom_bar(stat = "identity") + 
  labs(title = "Number of Delays by Service Type", x = "Service Type", y = "Number of Delays", fill = "Service Type") +  # Label the axes
  theme_minimal()  # Apply a minimal theme
```

## Most Delayed during Weekday
Define Rush Hours

Present Results During Rush Hours & at Off-Peak Hours

```{r}
#| echo: false
#| message: false

#filter out delays that occurred on weekdays
weekday_bus_delay <- bus_delay |>
  filter(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
weekday_streetcar_delay <- streetcar_delay |>
  filter(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
weekday_subway_delay <- subway_delay |>
  filter(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))

#filter out delays that occurred on weekends
weekend_bus_delay <- bus_delay |>
  filter(day %in% c("Saturday", "Sunday"))
weekend_streetcar_delay <- streetcar_delay |>
  filter(day %in% c("Saturday", "Sunday"))
weekend_subway_delay <- subway_delay |>
  filter(day %in% c("Saturday", "Sunday"))
```


```{r}
#| label: fig-weekday
#| fig-cap: Visualization to show the most delayed services on weekdays rush hours and off-peak hours
#| echo: false
#| message: false


```


```{r}
#| label: fig-weekend
#| fig-cap: Visualization to show the most delayed services on weekdays rush hours
#| echo: false
#| message: false


```

## Most Delayed during Weekends

```{r}
#| label: fig-informed
#| fig-cap: Visualization to show correlation between number of callers given information on homelessness prevention (represented by the dark green line) and Homeless Death Counts (represented by the blue line)
#| echo: false
#| message: false



```

## Most Popular Delay Reasons
- 1 table for all three services
- separate into rush & off-peak hours

## Most Popular Delayed Vehicles
- 1 table for all three services
- separate into rush & off-peak hours


# Discussion {#sec-discussion}

## TTC Service Prices {#sec-price}
- one-fare

## Correlation between Delay and Direction of Travel
- Financial District
- Midtown Direction

## TTC Service Amenities
- The PATH (for financial district)
- AC in Summer & Winter

## Errors in analysis, Limitations and Next Steps





# Conclusion {#sec-conclusion}






\newpage


# References


